---
title: "Data Exploration of Montgomery County, Maryland Crash Data"
author: "Carlos, Sean, and Sathyan"
output:
  html_document:
    df_print: paged
---
**Introduction**

---
**Data Acquisition**

We downloaded the data from <https://data.montgomerycountymd.gov/Public-Safety/Crash-Reporting-Drivers-Data/mmzv-x632> on March 13, 2018 which contained data updated through March 8, 2018 "on motor vehicle operators (drivers) involved in traffic collisions occurring on county and local roadways."

```{r echo=T}
	# na.strings parameter sets CELLS of blank, space, NA, and N/A to the NA reserved value
	dat <- read.csv("Crash_Reporting_-_Drivers_Data.csv",stringsAsFactors = F
		, na.strings = c(""," ","NA","N/A"))
	# create backup
	dat_bkup <- dat
```

---
**Data Cleanup**

Some issues involved
```{r echo=T}
	naCols <- names(dat)[grep("N/A",dat)]
```
* The data frame has many cells with "N/A" included among a string of values in columns: **
`r  naCols ` **
* The date/time field data type is actually a  `r class(dat$Crash.Date.Time) ` type

```{r echo=T}
# strip N/A string from lists of values
	dat[,naCols] <- apply(dat[,naCols], 2, function(x) gsub("N/A, ","",x))
	dat[,naCols] <- apply(dat[,naCols], 2, function(x) gsub(", N/A","",x))
# create a timestamp from the Crash Date Time string
	dat$timesStamp <- as.POSIXct(strptime(dat$Crash.Date.Time, tz = "America/New_York"
		,format = "%m/%d/%Y %I:%M:%S %p"))
```
---
**Data Exploration**

The data has `r nrow(dat)` rows and `r ncol(dat)` columns.



```{r echo=T}
str(dat)
head(dat)
sum(!complete.cases(dat))
names(dat)

sort(table(dat$Route.Type), decreasing = T)
sort(table(dat$Collision.Type),decreasing = T)
table(dat$Collision.Type,dat$Route.Type)
prop.table(table(dat$Collision.Type,dat$Route.Type),2)
par(mar=c(3,15,3,3))
barplot(table(dat[dat$Driver.Substance.Abuse == "ALCOHOL CONTRIBUTED",]$Collision.Type),horiz = T,las=2)
dat_dark <- subset(dat,dat$Light %in% c("DARK -- UNKNOWN LIGHTING","DARK LIGHTS ON","DARK NO LIGHTS"))
prop.table(table(dat_dark$Collision.Type,dat_dark$Injury.Severity),2)
prop.table(table(dat$Collision.Type,dat$Injury.Severity),1)

```


```{r echo=T}
# cleanup
#rm(naCols)
#rm(dat,dat_bkup,dat_dark)
```

